---
title: "Project 2"
author: "Matthew Herndon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project_2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This package was developed for Project 2 for MATH5793, which calls for the creation of an S3 class for performing multivariate techniques on datasets. This package specifically aims to validate whether or not each variable within a given 2D dataset is distributed approximately normally. The package may be installed as follows:

```{r, eval=FALSE}
devtools::install_github('matt-m-herndon/MATH5793_project_2')
```

# Theory

For a given one dimensional dataset, whether or not a 


# Classes

Two classes are defined in the package. The simplest class, **normal_assessment**, aims to answer the question "is a given 1D dataset normal?". The more complex class, **normal**, aims to answer the same questions about each variable within 2D datasets while also attempting to find transformations to normality when variables aren't well behaved. 

## normal_assessment

At its core, this class aims only to determine if a given **1-D** dataset is normal. To do this, a version of the QQ Correlation test (wherein a linear model is fit to the QQ data and correlated against the observations) is implemented.

Tp

This class contains information relating to the results,  

  obj = list(
    x=x,
    isnormal=correlation>=critical_point,
    model=model,
    sorted_samples=sorted,
    quartile_samples=q,
    model_samples=model_samples,
    correlation=correlation,
    critical_point=critical_point,
    significance=significance
  )
  
S3 methods are defined for **print**, **plot**, and **summary** which provide the user   

this defines methods for describing a one dimenbsio


The following show example outputs of the S3 methods defined for this class.

### Summary / Print

```{r}
library(project2normal)
# Grab a single variable from a single species from the iris dataset
assessment = normal_assessment(iris[1:25,1,])
summary(assessment)
```

### Plot

```{r}
plot(assessment)
```


## normal

Put sim

The following show example outputs of the S3 methods defined for this class.

### Summary / Print

```{r}
library(project2normal)
# Grab a single species from the iris dataset
norm_collection = normal(iris[1:25,,])
# All variables are summarized
summary(norm_collection)
```

### Plot

```{r}
# All variables are plotted
plot(norm_collection)
```

# Other features


